Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-05-13 ebayes_rnaseq_heterosis.tex
to eBayes

We present a hierarchical model for the gene-specific mean parameters and overdispersion parameters. The parental averages and overdispersion parameters are assumed to follow a normal distribution, i.e.
\[ 
%\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \quad \mbox{and}\quad 
\psi_g \stackrel{ind}{\sim} N(0,\sigma_\psi^2). 
\]
To allow for more shrinkage with the half-parental and hybrid differences, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e. 
\[ 
\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad \mbox{and} \qquad
\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
\]
where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$. We assume independence amongst the normalization factors, parental averages, half parental differences, hybrid differences, and overdispersion parameters.

We assume improper uniform priors for $\eta_\phi$, $\eta_\alpha$, and $\eta_\delta$ and independent half-cauchy priors for all standard deviations \citep{gelman2006prior}.

@@

We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$, and $\sigma=(\sigma_\gamma,\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$. We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on this MAP estimate. Conditional on the hyperparameters, the gene-specific parameters are independent, i.e., 
\begin{equation}
\begin{split}
p\left(\theta\left|y,\hat{\pi}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
%&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] p(\theta_g|\hat{\eta},\hat{\sigma}) \\
\propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
&\times 
\left. \vphantom{\prod_{v=1}^V}
N(\phi_g;\hat{\eta}_\phi, \hat{\sigma}_\phi^2)
La(\alpha_g;\hat{\eta}_\alpha, \hat{\sigma}_\alpha)
La(\delta_g;\hat{\eta}_\delta, \hat{\sigma}_\delta)
N(\psi_g;\hat{\eta}_\psi, \hat{\sigma}_\psi^2) 
\right]
\label{e:condind}
\end{split}
\end{equation}
Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.

>>>>

We present a hierarchical model for the gene-specific mean parameters and overdispersion parameters. The parental averages and overdispersion parameters are assumed to follow a normal distribution, i.e.
\[ 
%\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \quad \mbox{and}\quad 
\psi_g \stackrel{ind}{\sim} N(0,\sigma_\psi^2). 
\]
To allow for more shrinkage with the half-parental and hybrid differences, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e. 
\[ 
\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad \mbox{and} \qquad
\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
\]
where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$. We assume independence amongst the parental averages, half parental differences, hybrid differences, and overdispersion parameters.

%We assume improper uniform priors for $\eta_\phi$, $\eta_\alpha$, and $\eta_\delta$ and independent half-cauchy priors for all standard deviations \citep{gelman2006prior}.

@@

We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$, and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$. We obtain estimates for the hyperparameters and then base gene-specific inference on the posterior conditional on these estimates.

To obtain estimates of the hyperparameters, we use \edgeR to provide estimates of gene-specific parameters, $\hat{\theta}$ and normalization factors, $\hat{\gamma}$.

Conditional on the hyperparameters, the gene-specific parameters are independent, i.e., 
\begin{equation}
\begin{split}
p\left(\theta\left|y,\hat{\pi}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
%&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] p(\theta_g|\hat{\eta},\hat{\sigma}) \\
\propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
&\times 
\left. \vphantom{\prod_{v=1}^V}
N(\phi_g;\hat{\eta}_\phi, \hat{\sigma}_\phi^2)
La(\alpha_g;\hat{\eta}_\alpha, \hat{\sigma}_\alpha)
La(\delta_g;\hat{\eta}_\delta, \hat{\sigma}_\delta)
N(\psi_g;\hat{\eta}_\psi, \hat{\sigma}_\psi^2) 
\right]
\label{e:condind}
\end{split}
\end{equation}
Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
